---
title: "Reproducible Research Assignment 1"
author: "Yugandhar"
date: "Wednesday, July 08, 2015"
output: html_document
---

# This is a markdown file for the 1st assigment in "Reproducible Research"
# Data (activity.csv) was downloaded from the coursera webpage link. 

# Data is read into a data frame using read.csv

```{r echo=TRUE}
data <- read.csv("activity.csv")
```

## Remove all NA entries in the "steps"column into a new file.

``` {r echo=TRUE}
no_na <- data[!is.na(data$steps),]
```

# What is mean total number of steps taken per day?
### Use aggregate to sum number of steps by day and plot histogram
### mean and median values of number of steps per day superimposed
### on the graph with red line for mean and green dotted line for median

```{r echo=TRUE}
datstep <- aggregate(steps~date,no_na,sum)
hist(datstep$steps, ylab="days",xlab="steps", breaks = 10,main= "steps per day")
abline(v=mean(datstep$steps),col ="red",lwd =2)
abline(v=median(datstep$steps),col = "green",lwd=3,lty=2)
```
## ..
### values for mean and median number of steps taken in a day
```{r echo=TRUE}
mn <-mean(datstep$steps)
mn

md <- median(datstep$steps)
md


```

#What is the average daily activity pattern?
### Time series plot averaged across all days for the given 5min interval

```{r echo=TRUE}

stepint <- aggregate(steps~interval,no_na,mean)
plot(steps ~ interval,stepint,type="l",xlab= "minutes" ,ylab = "Average steps/day",main = "Avg Steps in a day")
intmaxstep <- stepint[stepint$steps== max(stepint$steps),]

intmaxstep$interval
```
### is the interval with maximum steps

# Imputing missing values with average steps with mean for that interval
### missing values were imputed with the mean value for a given interval

```{r echo=TRUE}
NA_num <- sum(is.na(data$steps))
NA_num

data1 <- transform(data, steps = ifelse(is.na(data$steps), stepint$steps[match(data$interval, stepint$interval)], data$steps))

```

# Histogram of imputed dataset
```{r echo=TRUE}

impdata <- aggregate(steps~date,data1,sum)
hist(impdata$steps,ylab="days",xlab="totalsteps", breaks = 10,main= "steps per day(imputed data")

imp_mn <-mean(impdata$steps)
mn

imp_md <- median(impdata$steps)
md

## There is no significant difference from imputing missing values to the data
## there is no difference in the mean value and 1.18 step difference in median

dif_in_mean = (imp_mn - mn)
dif_in_mean

dif_in_median = (imp_md - md)
dif_in_median

```

# Difference in activity pattern between weekdays and weekends

```{r echo=TRUE}

library(dplyr)

data2 <- data1
data2["wkdays"] <- NA
data2$wkdays <- weekdays(as.POSIXlt(data2$date))
data3<- data2 %>% mutate(weekend = ifelse(wkdays %in% c("Saturday", "Sunday"), "Weekend", "Weekday"))

par(mfrow = c(1, 2))
for (type in c("Weekend", "Weekday")) {
    steps.type <- aggregate(steps ~ interval, data = data3, subset = data3$weekend == 
        type, FUN = mean)
    plot(steps.type, type = "l", main = type)
}
```
### ..

### There is a difference in the activity patterns between weekdays and weekends as visible in the graph. During weekends, the activity intensity is for longer duration and it also sustained through the day. During weekdays there is an intense burst of activity in the early hours and never reaches the same later in the day.